### 2020-03-05 Common/chapters/literature_review.tex
Johan and Julien: Thesis layout updated until transfer learning :)

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}As stated by Gao et al., "Prostate  cancer  is  the  most  common  malignancies  among  men  and  remains  a  second  leading  cause to deaths in men globally. It was predicted that there would be 1.7 million new cancer cases by 2030. The early detection and diagnosis of prostate cancer can help to survive nine out of 10 men for the last five years"\cite{41}. Therefore, researchers proposed many different models to achieve good performance in prostate cancer detection. All the research works mentioned below are based on the SPIE-AAPM-NCI PROSTATEx Challenge dataset. This dataset is composed of multiparametric MRIs (T2W, DWI, ADC, DCE, PD, Ktrans) for a total of 204 training patients and 208 challenge patients (see Section \ref{sec:prostatex_dataset_description}).

Song et al.~\cite{07} presented a DCNN method to detect prostate cancer on multiparametric MRIs. Their data processing approach kept T2W, DWI and ADC grayscale images only. After resampling each image to the same resolution, T2W, DWI and ADC images were first cropped ($65$x$65$px patch) with the lesion in the center and stacked per patient, resulting in images containing three grayscale channels. Thanks to this method, the same lesion is visible in the same area over the three channels. This increases the probability of detecting a cancer by ensuring a good visibilty for each lesion, since the latter is not necesarily as visible with each parameter. Images were then normalized based on the Z-score per patient and per sequence (T2W, DWI, ADC), i.e. by subtracting the mean before dividing by the standard deviation. The data was split into a training ($80\%$), validation ($10\%$) and test set ($10\%$). The training (undefined number of times), validation (undefined number of times) and test images (11x) were augmented using -20$^\circ$ to 20$^\circ$ rotations, horizontal flipping, vertical sliding of less than 2 pixels and stretching by a factor between 0.9 and 1.1. Most of the processing techniques are reproductible, apart from the manual lesion contouring and labelling performed by a radiologist. 
Their model is a modified version of the well-known VGG16 model, including the addition of $1$x$1$ convolutions and dropout layers after each max pooling layer, and the use of the ELU activation function (see figure \ref{fig:paper_model}). The evaluation method for each patient and finding made an average of the 11 predictions resulting from the 11-time augmentation of the test set. The best results were obtained by using DWI images only with the highest b-value, reaching an AUC of $0.944$ with a 95$\%$ confidence interval ($0.876$-$0.994$). However, this model was not tested on the official PROSTATEx challenge images, which is an interesting benchmark to evaluate how well a model generalizes.

Liu et al.~\cite{46} created another architecture called XMasNet which was tested on the actual PROSTATEx challenge, achieving the second best performance at the time with an AUC of $0.84$. The AUC on their validation set reached $0.92$. Regarding data processing, their approach stacked different combinations of the available sequences as the three channels instead of defining a single combination like Song et al.: DWI-ADC-Ktrans, DWI-ADC-T2W, ADC-Ktrans-T2W and DWI-Ktrans-T2W. The data augmentation process differs in that the images are rotated in 3D, each lesion being sliced at 7 different orientations. These 2-dimensional slices were then augmented using rotation, shearing and translation of 1px, resulting in 207144 training samples. Both validation and testing test were also augmented in the same manner. This whole process allows to include 3-dimensional information in 2-dimensional images. The method used ensemble learning which combined different models to reach the best performance possible.

Mehrtash et al.~\cite{01} used a different approach. First of all, the input was fed to three separated parts of the model, each one responsible for a specific sequence among ADC, maximum b-value DWI and Ktrans. Then, each of these feature extractors' outputs were merged into a common decision maker. Furthermore, 3-dimensional convolutions instead of 2-dimensional ones were performed. In fact, 3-dimensional patches centered on the lesion were cropped. Augmentation including translation and flipping was used in order to balance the dataset. Apart from these differences, other minor differences such as normalizing the images within the range $[0,1]$ exist compared to the previous papers. These tricks allowed their model to achieve an AUC of $0.80$ on the PROSTATEx challenge. To make predictions, five different models were used, averaging the predictions of the four best models.

Armato et al~\cite{42} summarized the results obtained by all teams that took part in the PROSTATEx challenge in 2017. This challenge was split into two separate tasks. The first one was devoted to the diagnosis of prostate lesions (classification) whereas the other was about the segmentation and the determination of Gleason Grade Group. Thirty-two groups submitted results to the first challenge out of a total of 71 different methods (each group was allowed to submit up to three methods for evaluation). The article indicates that "most, but not all, methods outperformed random guessing (AUC=0.5)"~\cite{41}. The best performing method obtained an AUC value of 0.87 (standard error 0.027) and the next three methods all achieved AUC values of 0.84 (with standard errors of 0.036, 0.032, and 0.032). The complete results are illustrated on figure \ref{fig:challenge_all_results}. The median AUC on the challenge is approximately $0.68$.
\begin{figure}[!h]
\centering
\includegraphics[width=1\textwidth, keepaspectratio=true]{./figures/challenge_all_results.png}
\caption{AUC values achieved by the 71 methods that participated in the PROSTATEx Challenge~\cite{42}}
\label{fig:challenge_all_results}
\end{figure}

\section{Lung/Lung CT Challenge}

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan \& Julien}}Gao et al. claimed that "lung cancer is one of the most frequent and leading causes of death all over the world. It was reported that there were approximately $1.8$x$10^6$ new cases of lung cancer globally in 2012. Early detection of lung cancer, which is typically viewed in the form of lung nodules, is an efficient way to improve the survival rate"~\cite{41}. The articles cited below make use of the SPIE-AAPM Lung CT Challenge dataset, which is composed of CT scans devoted to the classification of lung nodules (see \ref{sec:lungCTChallenge} for further details).

Cengil et al.~\cite{02} built a fairly simple convolutional neural network to classify the images of the Lung CT Challenge dataset. The model takes 4-dimensional data as input (depth, height, width and channels) and performs 3D convolutions on it. The model consists of an input layer, five layers of 3D convolutions (the first is associated with a RELU activation function and pooling, the last with nothing, and the others with pooling) and a fully connected layer at the end. Regarding the model evaluation, authors announce an accuracy of 0.7 on their test set which contains 30 findings.

Armato et al.~\cite{12} described the LUNGx Challenge and the overall results. This challenge consisted in classifying the lung nodules as benign or malignant among a training set of 10 scans and a test set of 60 scans. Since the training set was extremely small, training the model on other datasets was allowed. The article describes the results of the proposed methods and compares them with the performance of six qualified radiologists on the same task. The article reports that "ten groups applied their own methods to 73 lung nodules (37 benign and 36 malignant) that were selected to achieve approximate size matching between the two cohorts. Area under the receiver operating characteristic curve (AUC) values for these methods ranged from 0.50 to 0.68. Only three methods performed statistically better than random guessing. The radiologists’ AUC values ranged from 0.70 to 0.85. Three radiologists performed statistically better than the best-performing computer method."~\cite{12}. Figure \ref{fig:LUNGx_challenge_all_results} gives an overview of the results.

@@

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan}}Brain cancer is another major type of cancer. According to American Society of Clinical Oncology, "brain and other nervous system cancer is the 10th leading cause of death for men and women. It is estimated that 17760 adults (9910 men and 7850 women) will die from primary cancerous brain and central nervous system tumors this year"~\cite{43}. The following research papers are based on the Kaggle "Brain MRI Images for Brain Tumor Detection" dataset. This dataset, unlike the ones for the other body parts, does not come from a certified medical authority but from the Kaggle website \cite{45}. However, since it is the only dataset available for brain tumor classification, some publications used it. Further details about this dataset can be found in section \ref{sec:kaggleBrain}.

Saxena et al.~\cite{31} implemented three convolution neural networks to classify the brain tumors coming from the Kaggle "Brain MRI Images for Brain Tumor Detection" dataset. Their processing method used a cropping technique which removed extra black margin around the skull. Each border of the image merged with a part of the skull. Since the images come from different sources, their resolution vary quite a lot. Therefore, the authors resized them to $224$x$224$x$3$. Moreover, data was augmented by rotation, vertical shifting and horizontal shifting. As the cropping was performed before augmenting the images, small parts of the brain were outside of the augmented images due to rotation and shifting. The data was split into a training set, a validation set and a test set. Regarding the models, authors implemented three of them (a Resnet-50, a VGG-16 and an Inception-V3) in order to compare their performance. The best results on the test set were achieved by the Resnet-50 (AUC of $0.95$ and accuracy of $0.95$). The VGG-16 was close (AUC of $0.90$ and accuracy of $0.90$), whereas the Inception-V3 didn't perform well (AUC of $0.55$ and accuracy of $0.55$).

%Hanwat et al.~\cite{32} implemented a convolutional neural network to classify brain tumors of the Kaggle dataset. Their processing method included skull masking, i.e. the removal of every non-brain tissue from the image. According to the authors, this technique improves performance quite a lot. ---> Check their methods (graphs look special)

Habib Mohamed Ali~\cite{04} proposed his own convolutional neural network to classify the images of the Kaggle Brain dataset. First, the data was augmented. Then, it was cropped thanks to OpenCV so that the resulting images only contained the brain itself. Afterwards, they were resized and normalized in order to scale pixel values to the range $[0,1]$. Once the processing part was over, the data was split into a training set ($70\%$), a validation set ($15\%$) and a test set ($15\%$). Regarding the neural network structure, the model is simple. It is composed of only one convolutional layer with a batch normalization layer and ReLU activation function, followed by two max-pooling layers and a dense layer. This model achieved an accuracy of $89\%$ on the test set.

>>>>

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}As stated by Gao et al., "Prostate  cancer  is  the  most  common  malignancies  among  men  and  remains  a  second  leading  cause to deaths in men globally. It was predicted that there would be 1.7 million new cancer cases by 2030. The early detection and diagnosis of prostate cancer can help to survive nine out of 10 men for the last five years"\cite{41}. Therefore, researchers proposed many different models to achieve good performance in prostate cancer detection. All the research works mentioned below are based on the SPIE-AAPM-NCI PROSTATEx Challenge dataset. This dataset is composed of multiparametric MRIs (T2W, DWI, ADC, DCE, PD, Ktrans) for a total of 204 training patients and 208 challenge patients (see Section \ref{sec:prostatex_dataset_description}).\\ \\
Song et al.~\cite{07} presented a DCNN method to detect prostate cancer on multiparametric MRIs. Their data processing approach kept T2W, DWI and ADC grayscale images only. After resampling each image to the same resolution, T2W, DWI and ADC images were first cropped ($65$x$65$px patch) with the lesion in the center and stacked per patient, resulting in images containing three grayscale channels. Thanks to this method, the same lesion is visible in the same area over the three channels. This increases the probability of detecting a cancer by ensuring a good visibilty for each lesion, since the latter is not necesarily as visible with each parameter. Images were then normalized based on the Z-score per patient and per sequence (T2W, DWI, ADC), i.e. by subtracting the mean before dividing by the standard deviation. The data was split into a training ($80\%$), validation ($10\%$) and test set ($10\%$). The training (undefined number of times), validation (undefined number of times) and test images (11x) were augmented using -20$^\circ$ to 20$^\circ$ rotations, horizontal flipping, vertical sliding of less than 2 pixels and stretching by a factor between 0.9 and 1.1. Most of the processing techniques are reproductible, apart from the manual lesion contouring and labelling performed by a radiologist. 
Their model is a modified version of the well-known VGG16 model, including the addition of $1$x$1$ convolutions and dropout layers after each max pooling layer, and the use of the ELU activation function (see figure \ref{fig:paper_model}). The evaluation method for each patient and finding made an average of the 11 predictions resulting from the 11-time augmentation of the test set. The best results were obtained by using DWI images only with the highest b-value, reaching an AUC of $0.944$ with a 95$\%$ confidence interval ($0.876$-$0.994$). However, this model was not tested on the official PROSTATEx challenge images, which is an interesting benchmark to evaluate how well a model generalizes.\\ \\
Liu et al.~\cite{46} created another architecture called XMasNet which was tested on the actual PROSTATEx challenge, achieving the second best performance at the time with an AUC of $0.84$. The AUC on their validation set reached $0.92$. Regarding data processing, their approach stacked different combinations of the available sequences as the three channels instead of defining a single combination like Song et al.: DWI-ADC-Ktrans, DWI-ADC-T2W, ADC-Ktrans-T2W and DWI-Ktrans-T2W. The data augmentation process differs in that the images are rotated in 3D, each lesion being sliced at 7 different orientations. These 2-dimensional slices were then augmented using rotation, shearing and translation of 1px, resulting in 207144 training samples. Both validation and testing test were also augmented in the same manner. This whole process allows to include 3-dimensional information in 2-dimensional images. The method used ensemble learning which combined different models to reach the best performance possible.\\ \\
Mehrtash et al.~\cite{01} used a different approach. First of all, the input was fed to three separated parts of the model, each one responsible for a specific sequence among ADC, maximum b-value DWI and Ktrans. Then, each of these feature extractors' outputs were merged into a common decision maker. Furthermore, 3-dimensional convolutions instead of 2-dimensional ones were performed. In fact, 3-dimensional patches centered on the lesion were cropped. Augmentation including translation and flipping was used in order to balance the dataset. Apart from these differences, other minor differences such as normalizing the images within the range $[0,1]$ exist compared to the previous papers. These tricks allowed their model to achieve an AUC of $0.80$ on the PROSTATEx challenge. To make predictions, five different models were used, averaging the predictions of the four best models.\\ \\
Armato et al~\cite{42} summarized the results obtained by all teams that took part in the PROSTATEx challenge in 2017. This challenge was split into two separate tasks. The first one was devoted to the diagnosis of prostate lesions (classification) whereas the other was about the segmentation and the determination of Gleason Grade Group. Thirty-two groups submitted results to the first challenge out of a total of 71 different methods (each group was allowed to submit up to three methods for evaluation). The article indicates that "most, but not all, methods outperformed random guessing (AUC=0.5)"~\cite{41}. The best performing method obtained an AUC value of 0.87 (standard error 0.027) and the next three methods all achieved AUC values of 0.84 (with standard errors of 0.036, 0.032, and 0.032). The complete results are illustrated on figure \ref{fig:challenge_all_results}. The median AUC on the challenge is approximately $0.68$.
\begin{figure}[!h]
\centering
\includegraphics[width=1\textwidth, keepaspectratio=true]{./figures/challenge_all_results.png}
\caption{AUC values achieved by the 71 methods that participated in the PROSTATEx Challenge~\cite{42}}
\label{fig:challenge_all_results}
\end{figure}

\section{Lung/Lung CT Challenge}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan \& Julien}}Gao et al. claimed that "lung cancer is one of the most frequent and leading causes of death all over the world. It was reported that there were approximately $1.8$x$10^6$ new cases of lung cancer globally in 2012. Early detection of lung cancer, which is typically viewed in the form of lung nodules, is an efficient way to improve the survival rate"~\cite{41}. The articles cited below make use of the SPIE-AAPM Lung CT Challenge dataset, which is composed of CT scans devoted to the classification of lung nodules (see \ref{sec:lungCTChallenge} for further details).\\ \\
Cengil et al.~\cite{02} built a fairly simple convolutional neural network to classify the images of the Lung CT Challenge dataset. The model takes 4-dimensional data as input (depth, height, width and channels) and performs 3D convolutions on it. The model consists of an input layer, five layers of 3D convolutions (the first is associated with a RELU activation function and pooling, the last with nothing, and the others with pooling) and a fully connected layer at the end. Regarding the model evaluation, authors announce an accuracy of 0.7 on their test set which contains 30 findings.\\ \\
Armato et al.~\cite{12} described the LUNGx Challenge and the overall results. This challenge consisted in classifying the lung nodules as benign or malignant among a training set of 10 scans and a test set of 60 scans. Since the training set was extremely small, training the model on other datasets was allowed. The article describes the results of the proposed methods and compares them with the performance of six qualified radiologists on the same task. The article reports that "ten groups applied their own methods to 73 lung nodules (37 benign and 36 malignant) that were selected to achieve approximate size matching between the two cohorts. Area under the receiver operating characteristic curve (AUC) values for these methods ranged from 0.50 to 0.68. Only three methods performed statistically better than random guessing. The radiologists’ AUC values ranged from 0.70 to 0.85. Three radiologists performed statistically better than the best-performing computer method."~\cite{12}. Figure \ref{fig:LUNGx_challenge_all_results} gives an overview of the results.

@@

\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Johan}}Brain cancer is another major type of cancer. According to American Society of Clinical Oncology, "brain and other nervous system cancer is the 10th leading cause of death for men and women. It is estimated that 17760 adults (9910 men and 7850 women) will die from primary cancerous brain and central nervous system tumors this year"~\cite{43}. The following research papers are based on the Kaggle "Brain MRI Images for Brain Tumor Detection" dataset. This dataset, unlike the ones for the other body parts, does not come from a certified medical authority but from the Kaggle website \cite{45}. However, as it is the only dataset available for brain tumor classification, some publications used it. Further details can be found in section \ref{sec:kaggleBrain}.\\ \\
Saxena et al.~\cite{31} implemented three convolution neural networks to classify the brain tumors coming from the Kaggle "Brain MRI Images for Brain Tumor Detection" dataset. Their processing method used a cropping technique which removed extra black margin around the skull. Each border of the image merged with a part of the skull. Since the images come from different sources, their resolution vary quite a lot. Therefore, the authors resized them to $224$x$224$x$3$. Moreover, data was augmented by rotation, vertical shifting and horizontal shifting. As the cropping was performed before augmenting the images, small parts of the brain were outside of the augmented images due to rotation and shifting. The data was split into a training set, a validation set and a test set. Regarding the models, authors implemented three of them (a Resnet-50, a VGG-16 and an Inception-V3) in order to compare their performance. The best results on the test set were achieved by the Resnet-50 (AUC of $0.95$ and accuracy of $0.95$). The VGG-16 was close (AUC of $0.90$ and accuracy of $0.90$), whereas the Inception-V3 didn't perform well (AUC of $0.55$ and accuracy of $0.55$).\\ \\
%Hanwat et al.~\cite{32} implemented a convolutional neural network to classify brain tumors of the Kaggle dataset. Their processing method included skull masking, i.e. the removal of every non-brain tissue from the image. According to the authors, this technique improves performance quite a lot. ---> Check their methods (graphs look special)
Habib Mohamed Ali~\cite{04} proposed his own convolutional neural network to classify the images of the Kaggle Brain dataset. First, the data was augmented. Then, it was cropped thanks to OpenCV so that the resulting images only contained the brain itself. Afterwards, they were resized and normalized in order to scale pixel values to the range $[0,1]$. Once the processing part was over, the data was split into a training set ($70\%$), a validation set ($15\%$) and a test set ($15\%$). Regarding the neural network structure, the model is simple. It is composed of only one convolutional layer with a batch normalization layer and ReLU activation function, followed by two max-pooling layers and a dense layer. This model achieved an accuracy of $89\%$ on the test set.

### 2020-03-05 Common/chapters/medical.tex
Johan and Julien: Thesis layout updated until transfer learning :)

\section{Cancer}
\subsection{Basics}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}An accumulation of cells forming a mass is called a tumor. These tumors are detectable thanks to medical imaging (see section \ref{sec:medical_imaging}) and other symptoms. However, not every tumor is as dangerous as the other, as it can be benign (does not contain cancerous cells) or malignant (contains cancerous cells).

The term cancer refers to different phenomena which involve mutation, abnormal multiplication and spreading of cells. As stated by Hanahan et al. in "The Hallmarks of Cancer" \cite{19} and "The Hallmarks of Cancer: The Next Generation" \cite{20}, every malignant tumor acquires six different capabilities during its evolution: 
\begin{itemize}
	\item \textbf{"Sustaining proliferative signaling"}\\ Cancerous cells don't wait for the body's approval to grow and proliferate, contrary to normal cells. They become responsible for their own multiplication.
	\item \textbf{"Evading growth suppressors"}\\
The body sends signals to contain cell growth within a tissue. Cancerous cells are insensitive to these. 
	\item \textbf{"Activating invasion and metastasis"}\\
Metastases are cells whose role is to propagate to other parts of the body in order to colonize and create new tumors. 
	\item \textbf{"Enabling replicative immortality"}\\
Healthy cells replication is limited to a certain amount, which is not the case for cancerous cells. 
	\item \textbf{"Inducing angiogenesis"}
\\Angiogenesis is the process of creating new blood vessels. Tumors have an influence on angiogenesis around them, since they need vascularization to continue growing. 
	\item \textbf{"Resisting cell death"}
\\Apoptosis is the programmed death of cells, which is part of the continuous regeneration of every cell within a body. Cancerous cells survive this programmed death. 
\end{itemize}

@@

\section{Types of medical imaging}
\label{sec:medical_imaging}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Multiple types of medical imaging exist. The most commonly used to detect cancer are Magnetic Resonance Imaging (MRI), CT (Computed Tomography) scans
%, PET (Positron Emission Tomography) scans
and mammograms.

MRI relies on magnetic fields to provide a three-dimensional view of body parts, which allows to see the generated images as a volume. Different settings, usually called sequences, make the look of the output vary, as shown on figure \ref{fig:PROSTATEx-t2-adc-dwi}.
Unlike MRI, CT is based on X-rays instead of magnetic fields, but still provides a three-dimensional representation of a body part. Figure \ref{fig:visualize_lung_dcm} shows a lung CT scan.

@@

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}DICOM files can be viewed as containers of attributes, also called tags. The values of the pixels themselves are stored under the "Pixel Data" tag. Every single DICOM file usually represents a 2-dimensional image, which will form a 3-dimensional volume when put all together.

Other useful information such as the patient name and ID is directly stored within the DICOM files. This approach aims at linking each image to a specific person and event in order not to mix them up. Each DICOM file can be seen as part of a bigger dataset.

@@

\subsubsection{Data manipulation}
\label{sec:dicom_data_manipulation}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}CT and MRI machines, as well as monitors, differ from one manufacturer to the other and even from one model to the other. DICOM takes this problematic into account by providing specific tags that allow to display the exact same representation of the data, no matter the hardware used. Otherwise, physicians may struggle to detect anomalies because of color and exposition-related variations. 
Therefore, before displaying or converting an image to any format (such as png), pixel data must be normalized.

The procedure depends on the tags "Window Width" and "Window Center" (one always come with the other). These are used to represent a range of values corresponding to the pixel values in the data. For instance, a window center of 0 and a window width of 200 imply pixel values between -100 and 100.

If they are missing, a simple conversion is sufficient. The parameters used to convert the data are given by two tags: 
\begin{itemize}
	\item Bits allocated: the number of bits used to represent a single pixel (value: 1 or a multiple of 8)
	\item Samples per pixel: the number of channels for each pixel

\noindent Examples: 
\begin{itemize}
\item 1 bit, 1 sample: black and white
\item 8 bits, 1 sample: grayscale
\item 8 bits, 3 samples: RGB
\item 16 bits, 1 sample: grayscale

@@

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Unlike the ANALYZE format that used two files to store the metadata and the actual data, the NIfTI file format stores them in one single file “.nii” but keeps this split between the real data and the header for compatibility. This has the advantage to facilitate the use of the data and avoid storing the data without the meta-data. The NIfTI format can also be compressed/decompressed on-the-fly using the “deflate”  algorithm.

@@

\subsection{DICOM}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tools allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the Z-axis, displaying the next or previous slice.

>>>>

\section{Cancer}
\subsection{Basics}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}An accumulation of cells forming a mass is called a tumor. These tumors are detectable thanks to medical imaging (see section \ref{sec:medical_imaging}) and other symptoms. However, not every tumor is as dangerous as the other, as it can be benign (does not contain cancerous cells) or malignant (contains cancerous cells).\\
The term cancer refers to different phenomena which involve mutation, abnormal multiplication and spreading of cells. As stated by Hanahan et al. in "The Hallmarks of Cancer" \cite{19} and "The Hallmarks of Cancer: The Next Generation" \cite{20}, every malignant tumor acquires six different capabilities during its evolution: 
\begin{itemize}
	\item \textbf{"Sustaining proliferative signaling"}\\ Cancerous cells don't wait for the body's approval to grow and proliferate, contrary to normal cells. They become responsible for their own multiplication.
	\item \textbf{"Evading growth suppressors"}\\
The body sends signals to contain cell growth within a tissue. Cancerous cells are insensitive to these. 
	\item \textbf{"Activating invasion and metastasis"}\\
Metastases are cells whose role is to propagate to other parts of the body in order to colonize and create new tumors. 
	\item \textbf{"Enabling replicative immortality"}\\
Healthy cells replication is limited to a certain amount, which is not the case for cancerous cells. 
	\item \textbf{"Inducing angiogenesis"}
\\Angiogenesis is the process of creating new blood vessels. Tumors have an influence on angiogenesis around them, since they need vascularization to continue growing. 
	\item \textbf{"Resisting cell death"}
\\Apoptosis is the programmed death of cells, which is part of the continuous regeneration of every cell within a body. Cancerous cells survive this programmed death. 
\end{itemize}

@@

\section{Types of medical imaging}
\label{sec:medical_imaging}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Multiple types of medical imaging exist. The most commonly used to detect cancer are Magnetic Resonance Imaging (MRI), CT (Computed Tomography) scans and mammograms.\\
MRI relies on magnetic fields to provide a three-dimensional view of body parts, which allows to see the generated images as a volume. Different settings, usually called sequences, make the look of the output vary, as shown on figure \ref{fig:PROSTATEx-t2-adc-dwi}.
Unlike MRI, CT is based on X-rays instead of magnetic fields, but still provides a three-dimensional representation of a body part. Figure \ref{fig:visualize_lung_dcm} shows a lung CT scan.

@@

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}DICOM files can be viewed as containers of attributes, also called tags. The values of the pixels themselves are stored under the "Pixel Data" tag. Every single DICOM file usually represents a 2-dimensional image, which will form a 3-dimensional volume when put all together.\\
Other useful information such as the patient name and ID is directly stored within the DICOM files. This approach aims at linking each image to a specific person and event in order not to mix them up. Each DICOM file can be seen as part of a bigger dataset.

@@

\subsubsection{Data manipulation}
\label{sec:dicom_data_manipulation}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}CT and MRI machines, as well as monitors, differ from one manufacturer to the other and even from one model to the other. DICOM takes this problematic into account by providing specific tags that allow to display the exact same representation of the data, no matter the hardware used. Otherwise, physicians may struggle to detect anomalies because of color and exposition-related variations. 
Therefore, before displaying or converting an image to any format (such as png), pixel data must be normalized.\\
The procedure depends on the tags "Window Width" and "Window Center" (one always come with the other). These are used to represent a range of values corresponding to the pixel values in the data. For instance, a window center of 0 and a window width of 200 imply pixel values between -100 and 100.\\
If they are missing, a simple conversion is sufficient. The parameters used to convert the data are given by two tags: 
\begin{itemize}
	\item Bits allocated: the number of bits used to represent a single pixel (value: 1 or a multiple of 8)
	\item Samples per pixel: the number of channels for each pixel

\end{itemize}
\newpage
\noindent Examples: 
\begin{itemize}
\item 1 bit, 1 sample: black and white
\item 8 bits, 1 sample: grayscale
\item 8 bits, 3 samples: RGB
\item 16 bits, 1 sample: grayscale

@@

\subsection{Data format}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Unlike the ANALYZE format that used two files to store the metadata and the actual data, the NIfTI file format stores them in one single file “.nii” but keeps this split between the real data and the header for compatibility. This has the advantage to facilitate the use of the data and avoid storing the data without the meta-data. The NIfTI format can also be compressed/decompressed on-the-fly using the “deflate” algorithm. M. Winkler affirms that this is particularly useful as "it is very common for images to have large areas of solid background, or files describing masks and regions of interest containing just a few unique values that appear repeated many time"~\cite{52}.

@@

\subsection{DICOM}
\setlength{\marginparwidth}{3cm}\leavevmode \marginnote{\textbf{Julien}}Visualizing DICOM files is pretty straightforward since each file represents a single two-dimensional image. However, a lot of pixel transformations and normalizations have to be applied to obtain the desired result (see section \ref{sec:dicom_data_manipulation}), which may be a source of errors. Our tools allows to display a single DICOM file as well as a sequence of files if the function is fed with a directory. Users can then scroll through the Z-axis, displaying the next or previous slice.  Figure \ref{fig:visualize_lung_dcm} shows the visualization of a lung at three different Z-axis coordinates. This allows the user to simply have a complete overview of the body part he want to analyse.
